Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-08-31 review/Response to reviews.Rmd

We changed Fig 3 to be on the response scale as suggested (see our response to \@ref(R1-figs), but showing the raw data made the plots visually cluttered, in our opinion.
Another way we approached this comment is by excluding areas of the heatmaps in figures 4-6 that are far from observed data, equivalent to not drawing regression lines beyond the range of data.
We implemented this with the `exclude.too.far()` function in the `mgcv` package.
Displaying raw data on the heatmaps was too messy and not informative in a 2d plot.

##  {#R1-5}

> I suggest that a little more about the limitations of this modeling approach in the discussion would help set the context for interpreting the results and make the paper more robust.

The main limitations of this method are of data (see also reviewer 2's comment \@ref(R2-15)).
We have added some text to the discussion describing limitations found in previous simulation studies, and new unpublished evidence that spatial replication may be more valuable than long time series for detecting demographic effects of the environment.

@@

The reviewer is correct that recruitment is an important aspect of demography that could be effected by SPEI and fragmentation.
However, we believed the analyses in this paper were complicated enough to not include additional analyses.
We intend to follow up this paper with analyses of recruitment and integrating all vital rate models with an integral projection model.
We simply felt doing all of this (describing and interpreting DLNMs, modeling recruitment, and integrating all into an IPM) would be too much for a single paper.

##  {#R1-7}

> I had a hard time linking these statements in the results to the figures.
>
> L272--273---How do you see disparities in plant size here?
> I thought this was about growth?

Models which predict size in year t+1 as a function of size in year t, such as ours, are often described as growth functions because they can be used to project change in size (i.e. growth).
However, we agree that it would be more clear for this section heading to be "size" rather than "growth".

@@

> L277--278---How does this show the effect of size on growth is greatest for mid-size plants?
> More generally, I got a little confused with the flipping back and forth between growth & plant size in the results and discussion.
> More precise language would help.

We have edited the text to omit "growth", and hopefully clarify: "...the impact of plant size on size in the following year was greatest for mid-sized plants in continuous forest."

@@

Our point was actually that SPEI had less of an impact on flowering probability in fragments.
We edited this sentence to read: "SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure 6b)."

##  {#R1-8}

> L442-443 -- What is a mechanism for why plants might flower more after drought?
> (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.

We are a little hesitant to speculate too much here, as not much is known about the optimal microclimate conditions for flowering in *H. acuminata*, but we have revised this paragraph (see also our response to @ref(R2-17). It now includes the following possible explanation: *"Drought conditions could be favorable for H. accuminata flowering due to increased temperatures or decreased cloudiness associated with droughts*[@pauCloudsTemperatureDrive2013]*,* *and the effects could be delayed due to the development time of inflorescences."*

@@

See our response to \@ref(R2-7).

##  {#R2-13}

> Line 412: I think it would be very informative if you were to expand on this and explain how exactly your method differs from what others have used to detect delayed effects.

As reviewer 3 pointed out (\@ref(R3-11)), our extension to previous approaches is the use of a 2-dimensional spline to allow for fitting non-linear responses to the environment.
We have edited this sentence accordingly

@@

See our response to \@ref(R2-7).

##  {#R2-15}

> You may also consider arguing that an 11-year long time series is not sufficient to detect delayed effects (see Tenhumberg et al. 2018)

We have added additional discussion on the limitations of DLNMs for detecting delayed effects to the discussion.
See our responses to \@ref(R1-5).

@@

We intended this paragraph to focus on one possible explanation for delayed effects---delays due to development times.
We believe that the bet-hedging argument was confusing for two reasons: 1) it assumed that drought (as we measured it with SPEI) is detrimental to *H. accuminata*, which is not consistently reflected in our results, and 2) it did not address longer lags than the development time of a flowering shoot.
We have therefore revised the example in this paragraph to focus on literature that shows positive effects of temperature and negative effects of cloudiness on flowering in tropical plants---weather conditions both associated with droughts.
See also our response to \@ref(R1-8).

##  {#R2-18}

> Line 448: Alternatively, drought may reduce nutrient availability such that after covering maintenance costs there is nothing left to allocate.

<!--# wouldn't this reduce growth?  Not what we saw in our results, so not sure how this is relevant? -->

@@

##  {#R2-19}

> Lines 451-453: I think herbivory is this is very different from drought.
> Drought reduced nutrient availability while herbivory removes some plant biomass triggering compensatory growth.

<!--# but we just said drought also removes tissue in the previous sentence.  Am I missing something? -->

@@

We agree with the reviewer that the exact timing of mortality within a season doesn't matter for population dynamics.
This is likely obvious to demographers who are used to working with yearly census data and finer scale climate data.
However, we think this paragraph is appropriate to follow up with the three before it---a delayed effect indicated by a DLNM plot like those in Figures 4, 5 and 6 could be due to 1) developmental delays, 2) buffering by below-ground storage tissues, 3) indirect effects, or 4) **none of the above.** We want to be sure to point out that in the case of survival especially, a "delayed effect" of SPEI 5 months in the past could simply mean that a plant died *during* a drought 5 months in the past (no actual delay between adverse conditions and death).
Importantly, this does not apply to flowering in our system, since flowering happens at a specific time of year (during the census), unlike mortality.
We have revised this paragraph to emphasize the above points.

##  {#R2-21}

> The discussion includes reasons for why we expect delayed effect in plants in general, but it does not relate the specifics of the significant delays to the life history of H.
> acuminata.
> For instance, why does too much rain reduce survival?
> Why is there an effect of SPEI during the dry season (18-20 months in the past but during the wet season 32-36 months in the past? Why do plants in fragments have a decrease survival when precipitation is high during dry and wet periods (1-18 months time lag)?
> Etc.

While we believe we did discuss *Heliconia*-specific mechanisms for delayed responses of flowering (L 434--438, now <!--# new line -->) and delayed effects on growth (L 445--450, now <!--# new line -->), the reviewer is correct that we had not previously discussed mechanisms behind mortality.

@@

##  {#R2-22}

> Lines 484-486: I don't think this study strongly support that plants do worse in fragments than CF.

Although the differences in plant size and flowering are relatively small, they are statistically significant (see Table 1 in revised manuscript), and we believe they are demographically significant over long time periods.
See our response to \@ref(R2-7).

@@

We moved the explanation of covariates and random effects earlier as suggested.

##  {#R3-4}

> Line 205 - I'm not sure "bimodal" is correct here -- a cubic function can have 2 local extrema, is that what you mean?

This is what we meant, and is no longer relevant as we have increased the number of knots.
See response to \@ref(R3-5).

@@

##  {#R3-5}

> Line 205-207 -- "number of knots not important ..." yes, but package mgcv includes tools for checking those assumptions, would be helpful to include residual analysis and the results of gam.check() in an appendix.

This comment was especially helpful and made us re-think limiting the number of "knots" to 3 in the SPEI dimension.
The term *k* is probably more accurately described as the "maximum basis complexity" or "maximum basis dimension".
We had thought that by limiting k to 3, we were simply limiting the complexity of the smooth in the SPEI dimension.
However, after taking the reviewer's suggestions we realized that there were other unexpected consequences.
We increased *k* for the SPEI dimension for all models and for other smooths for some models.
This resulted in some changes in the the shape of the crossbasis smooth and which areas of it were statistically significant.
We have updated the methods and results sections accordingly.\
\
We have also now included a supplemental document with model validation, including checks that *k* is adequately large
.

@@

See our response to \@ref(R1-figs).

##  {#R3-9}

> Line 260 -- you mention the CI of the model estimated coefficients, but these aren't provided anywhere that I could find.
> I think a table based on the summary of the two fitted models discussed here (and elsewhere too) should be added.
> And report the variance of the random effect.

We have now included these in Table 1.
See also our response to \@ref(R1-3).

@@

In the updated version of figure 4 (see response to \@ref(R3-5)), a smaller area of the plot is significantly different from the intercept, and heatmap area far from observed data is grayed out.
We believe these two things make it easier to see the outlines around significant values, but we are open to trying other modifications of the plot if the reviewers think it is still difficult to interpret in this way.

##  {#R3-11}

> Line 413 -- "our expansion of this approach ..." I'm not sure that the expansion is the penalty that prevents over fitting -- both Teller et al and Tenhumberg et al used the penalized spline function in mgcv.
> What IS an expansion is the non-linear cross basis function.
> Both Teller et al and Tenhumberg et al used (essentially) linear cross basis functions, so using dlnm is super-cool and interesting.

We thank the reviewer for the positive comment.
This is correct---our expansion was not the fitting of lagged effects with a spline, but the use of a 2 dimensional spline to *also* fit non-linear reaction norms.
We have edited this sentence accordingly.

### 2021-08-31 doc/paper.Rmd
Discuss pros and cons of DLNMs more. Closes #80.

Despite these insights, models of plant population dynamics rarely include the effects of environmental drivers [but see @williams2015; @tenhumbergTimelaggedEffectsWeather2018; @molowny-horasChangesNaturalDynamics2017).
This has largely been due to the challenge (both ecologically and statistically) of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.

@@

Finally, any analytical approach for assessing lagged effects on demography requires long-term data and/or data collected from many sites with independent weather (but similar climate and site conditions) [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021] <!--# add citation to Aldo's ESA talk -->.
<!--# possibly good place to mention limitations of interactions -->Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.
More generally, however, researchers need to consider how delayed responses to climate could influence the interpretation of data in studies where the organisms lifespan exceeds the study's duration.

>>>>

Despite these insights, models of plant population dynamics rarely include the effects of environmental drivers [but see @williams2015; @tenhumbergTimelaggedEffectsWeather2018; @molowny-horasChangesNaturalDynamics2017). This has largely been due to the challenge (both ecologically and statistically) of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons. To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021; @ogleQuantifyingEcologicalMemory2015].
Our extension of this approach allows for modeling of non-linear responses to the environment in addition to an unbiased way of identifying these delayed effects without overfitting.
Our results are consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.

@@

Finally, any analytical approach for assessing lagged effects on demography requires long-term data and/or data collected from many sites with independent weather (but similar climate and site conditions) [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021]. @tellerLinkingDemographyDrivers2016 used a simulation study to show that detecting lagged effects required 20--25 years of data, although this depended on effect size and the range of climate extremes in the data. Compagnoni et al. ( <!--# cite ms in prep?  cite ESA talk? --> ) showed that spatial replication can alleviate this requirement, and in fact in our study we were able to detects significant lagged effects with only 10 years of data, but multiple plots experiencing slightly different weather. Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.
More generally, however, researchers need to consider how delayed responses to climate could influence the interpretation of data in studies where the organisms lifespan exceeds the study's duration.
